Speeding up eQTL scans in the BXD population using GPU

### 2019-10-28 paper.tex
saving changes

%Why GPU?
Processing large data set has been a challenge for genome scan. 
We were blessed with Moore's Law for decades, but the Central Processing Unit (CPU) is reaching to its physical limit of transistors. 
Graphics Processing Unit (GPU), originally used as an image processing component of a computer, has shown some compelling results of accelerating computation in various fields.
General Purpose computing Graphics Processor (GPGPU) became popular in the early 2000s because of its ability to natively handel matrix and vector operations. 
Such power is very attractive to the scientific computing community. 
Zhang et al. \citep{zhang2015mixed} used GPU to simultaneously dissect various genetic effects with mixed linear model. 
Chapuis et all. \citep{chapuis2013graphics} utilizes GPU to offset heavy computation to deploy various ways for a more pricise detection of QTLs threashold. 
By using GPU-backed machine learning libraries like TensorFlow, Taylor-Weiner et al. \citep{taylor2019scaling} reimplemented QTL mapping  and Bayesian non-negative matrix factorization, achieving greater than 200 times speedup compared to CPU-based versions. 
The ease of using such library will push method-development for genomeic research.

@@

@@

		\end{tabularx}
		\footnotetext[1]{There are overhead of using GPU, such as data transfer and setting up context, the timing shown for CPU \& GPU here included all overhead for fairness. We ran the genome scan process 10 times and choose the median to remove randomness of each run, and the warm up time of GPU.}
		\label{tab:speedup-table}
	\end{tableminipage}
\end{table*}

\subsection{Benefit of using linear model}
Our method uses the linear model and simplified the genome scan process to basic matrix operations. 
The timing of our method is shown in Table \ref{tab:speedup-table}. 
The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
Getting the same result using the same dataset from R/qtl took about 36.5 seconds.

@@

The genome scan process includes the following steps:
\begin{itemize}
	\item Calculate standardized matrices ($G^{\ast}$, $Y^{\ast}$) for input matrices ($G$, $Y$)
	\item Get a correlation matrix ($R$) by multiplying the standardized matrices
	\item Calculate LOD scores 
\end{itemize}
Our profiling result shows that the second and third steps take up 90\% of the time and involve parallelizable matrix operations.  Hence, they are our candidates for GPU acceleration.
To make a fair comparison of the CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of the performance time.  That is, the former is rather slower than the latter regardless of precision for the data type.
We used a GPU profiler \textit{nvprof} \cite{nvprof}
to investigate what is the bottleneck of GPU. 
The result is, 98\% of the time is spent on data transfer from GPU to CPU (device to host).
As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
For BXD spleen dataset, Y' matrix is 17 MB, G matrix is 21 MB
while R matrix is about 4GB. 
Data offloading is the main bottleneck for our GPU implementation.

>>>>

%Why GPU?
Processing large data set has been a challenge for genome scan. 
We were blessed with Moore's Law for decades, but the Central Processing Unit (CPU) is reaching to its physical limit of transistors. 
Graphics Processing Unit (GPU), originally used as an image processing component of a computer, has shown some compelling results of accelerating computation in various fields.
General Purpose computing Graphics Processor (GPGPU) became popular in the early 2000s because of its ability to natively handel matrix and vector operations. 
Such power is very attractive to the scientific computing community. 
Zhang et al. \citep{zhang2015mixed} used GPU to simultaneously dissect various genetic effects with mixed linear model. 
Chapuis et all. \citep{chapuis2013graphics} utilizes GPU to offset heavy computation to deploy various ways for a more pricise detection of QTLs threashold. 
By using GPU-backed machine learning libraries like TensorFlow, Taylor-Weiner et al. \citep{taylor2019scaling} reimplemented QTL mapping  and Bayesian non-negative matrix factorization, achieving greater than 200 times speedup compared to CPU-based versions. 
The ease of using such library will push method-development for genomic research.

@@

%\subsection{Benefit of using linear model}
Our method uses the linear model and simplified the genome scan process to basic matrix operations. 
The timing of our method is shown in Table \ref{tab:speedup-table}. 
The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
Getting the same result using the same dataset from R/qtl took about 36.5 seconds.

@@

		\end{tabularx}

		\label{tab:speedup-table}
	\end{tableminipage}
\end{table*}

@@

The genome scan process includes the following steps:
\begin{itemize}
	\item Calculate standardized matrices ($G^{\ast}$, $Y^{\ast}$) for input matrices ($G$, $Y$)
	\item Get a correlation matrix ($R$) by multiplying the standardized matrices
	\item Calculate LOD scores 
\end{itemize}
Our profiling result shows that the second and third steps take up 90\% of the time and involve parallelizable matrix operations.  Hence, they are our candidates for GPU acceleration.
To make a fair comparison of the CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
There are overhead of using GPU, such as data transfer and setting up context, the timing shown for CPU \& GPU here included all overhead for fairness. We ran the genome scan process 10 times and choose the median to remove randomness of each run, and the warm up time of GPU.
The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of the performance time.  That is, the former is rather slower than the latter regardless of precision for the data type.
We used a GPU profiler \textit{nvprof} \cite{nvprof}
to investigate what is the bottleneck of GPU. 
The result is, 98\% of the time is spent on data transfer from GPU to CPU (device to host).
As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
For BXD spleen dataset, Y' matrix is 17 MB, G matrix is 21 MB
while R matrix is about 4GB. 
Data offloading is the main bottleneck for our GPU implementation.
